Higher-Order unification for free!

### 2024-04-15 paper/main.tex
sec 3.6, 3.6 and all 4 spellchecked

\noindent
So far the compiler is very basic, it does not really
enforce the terms passed to \hstep are in \wellb,
and indeed makes no use of the higher-order capabilities
of the meta language (all generated
variables have an empty scope).
Still, we can prove that \Uo is a good
``first order'' unification algorithm if the input
already happens to be in \wellb.
Later, when the compiler will enforce \cref{prop:w-enforcing}
and the proof will be adjusted to cover for thew cases.

\begin{lemma}[Properties of \Uo]\label{prop:uniffow} The following properties hold for \Uo:
\begin{gather}
  \unifcorrectX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}\\
  \unifcompleteX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}
\end{gather}
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as \Eo on ground terms.What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{} terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case after compilation we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

In the following sections we adopt this notation to talk about the compiler's
output: \foUnifPb is the input set of problems that compiled to
\hoUnifPb with memory mapping \mapStore and links \linkStore. For example:
%
\printAlll
  {{{p_1,p_2},{p_3,p_4}}}
  {{{t_1,t_2},{t_3,t_4}}}
  {{{X_1,A_1,{{x}}},{X_2,A_2,{{y}}}}}
  {{{\eta,\Gamma,a,b}}}

We index each sub-problem, sub-mapping, sub-link with its position
starting from $1$ and counting from left to right, top to bottom.
For example, $\hoUnifPb_2$ corresponds to the \Ho problem $t_3 \Ue t_4$.
%The compiled version of each $\foUnifPb_i$ is represented by $\hoUnifPb_i$.

@@

In order to make \Uo higher-order we need to
take care of terms in \maybebeta.
In the example below, we can see that the basic compilation given in 
the previous section is not able to make the \Ho unification problem succeeds.
%
% make test ONLY=7006 TEX=tex
\printAlll
  {{{\lambda x.(f\appsep (X\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{\lambda x.(f\appsep (A\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{X,A,0}}}
  {{}}
\todo{fix indentation}
The unification problem $\hoUnifPb_1$ fails while trying to unifying
$A \appsep x$ and $x$, equivalent to \elpiIn{app [uva A [], x]}
versus \elpiIn{x}.
In order to exploit the higher-order unification algorithm of the meta language,
we need to compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$
that is \elpiIn{uva A [x]}.

@@

We add the following rule before rule~\ref{rule:compapp}, where
\elpiIn{pattern-fragment} is a predicate checking is a list of
terms is a list of distinct names.

@@

\noindent
Note that compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
is made of bound variables and the hypothetical rule 
\cref{rule:complamh}
loaded by \elpiIn{comp-lam}
grants this property.

@@

  \begin{proof}[Proof sketch]
  If we look at the \Fo{} terms the is one more interesting case,
  namely \elpiIn{fapp[fuva X|W] ~\Uo~s} when \elpiIn{W}
  are distinct names compiled to $\vec{w}$. In this case
  the \Ho problem is $Y_{\vec{w}} \Ue t$ that succeeds with
   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{w}/\vec{y}]\}$ that in turn
   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{w}/\vec{y}]\}$.
   Thanks to \ref{clause:beta1}
   $(\lambda \vec{y}.s[\vec{w}/\vec{y}])~\vec{w} \Eo s$.
  \end{proof}

>>>>

\noindent
So far the compiler is very basic. It does not really
enforce that the terms passed to \hstep are in \wellb,
and indeed makes no use of the higher-order capabilities
of the meta language (all generated
variables have an empty scope).
Still, we can prove that \Uo is a good
``first-order'' unification algorithm if the input
already happens to be in \wellb.
Later, when the compiler will enforce \cref{prop:w-enforcing}
the proof will be adjusted to cover for the new cases.

\begin{lemma}[Properties of \Uo]\label{prop:uniffow} The following properties hold for \Uo:
\begin{gather}
  \unifcorrectX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}\\
  \unifcompleteX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}
\end{gather}
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting, \Ee is as strong as \Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed), we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{}, terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case, after compilation, we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

@@

In the following sections, we adopt this notation discuss the compiler's
output: \foUnifPb represents the input set of problems that compiled to
\hoUnifPb with memory mapping \mapStore and links \linkStore. For example:
%
\printAlll
  {{{p_1,p_2},{p_3,p_4}}}
  {{{t_1,t_2},{t_3,t_4}}}
  {{{X_1,A_1,{{x}}},{X_2,A_2,{{y}}}}}
  {{{\eta,\Gamma,a,b}}}

We index each sub-problem, sub-mapping, and sub-link with its position
starting from $1$ and counting from left to right, top to bottom.
For example, $\hoUnifPb_2$ corresponds to the \Ho problem $t_3 \Ue t_4$.
%The compiled version of each $\foUnifPb_i$ is represented by $\hoUnifPb_i$.

@@

In order to make \Uo higher-order, we need to
take care of terms in \maybebeta.
In the example below, we can see that the basic compilation given in 
the previous section is not able to make the \Ho unification problem succeeds.
%
% make test ONLY=7006 TEX=tex
\printAlll
  {{{\lambda x.(f\appsep (X\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{\lambda x.(f\appsep (A\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{X,A,0}}}
  {{}}
\todo{fix indentation}
The unification problem $\hoUnifPb_1$ fails while trying to unify
$A \appsep x$ and $x$, which is equivalent to \elpiIn{app [uva A [], x]}
versus \elpiIn{x}.
In order to exploit the higher-order unification algorithm of the meta language,
we need to compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$,
which is \elpiIn{uva A [x]}.

@@

We add the following rule before rule~\ref{rule:compapp}, where
\elpiIn{pattern-fragment} is a predicate checking if a list of
terms is a list of distinct names.

@@

\noindent
Note that compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
is made of bound variables, and the hypothetical rule 
\ref{rule:complamh}
loaded by \elpiIn{comp-lam}
grants this property.

@@

  \begin{proof}[Proof sketch]
  If we look at the \Fo{} terms, there is one more interesting case,
  namely \elpiIn{fapp[fuva X|W] ~\Uo~s} when \elpiIn{W}
  are distinct names compiled to $\vec{w}$. In this case
  the \Ho problem is $Y_{\vec{w}} \Ue t$ that succeeds with
   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{w}/\vec{y}]\}$, which in turn
   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{w}/\vec{y}]\}$.
   Thanks to \ref{clause:beta1}
   $(\lambda \vec{y}.s[\vec{w}/\vec{y}])~\vec{w} \Eo s$.
  \end{proof}
